Implementation of a stable, high-power optical lattice for quantum gas microscopy

### 2018-06-21 LatticeOptics/latticeOpticsv5/LatticeOpticsV5.tex

Variation in external conditions can contaminate large datasets if the prepared state sensitively depends on laser pointing.
To guard against such contamination, we have implemented an independent monitoring system to verify appropriate pointing of both lattice laser beams.
This monitoring system allows us to eliminate (via post-selection) experimental realizations during which the system was in an undesired state.
A beam-sampler (Thorlabs BSF10-C) reflects a small portion of the beam shortly before it reaches the atoms and images it onto a monitoring camera located in the plane corresponding to the atomic system, imaging a Gaussian beam corresponding to the optical lattice.
Naturally, the position of the atoms is dictated by the light field of both the incoming and retroreflected beams, while the monitoring camera only observes the position of the incoming beam.

@@

We first characterize the passive stability of the optical lattice along one dimension by monitoring its position over the course of 250 experimental realizations.
We decompose the lattice position into the transverse $x_t$ and the longitudinal $x_l$ components with respect to the optical axis.
Figure~\ref{fig:stability}(a) shows a histogram of displacements of atoms between subsequent experimental realizations, showing that the RMS displacement is $0.54(3)$ [$0.98(4)$] sites along the $x_t$ [$x_l$] direction.
The variation of the displacements is a geometric effect consistent with the mechanical assembly of the experiment~\cite{Huber2014}.
The distributions are centered within 1-$\sigma$ of zero along both directions: $-0.03(4)$ and $0.01(5)$ sites for transverse and longitudinal axes, respectively.
If the trap had followed a random walk with these step sizes, the optical lattice would have to be realigned frequently, since after only $100$ experimental shots, the trap would have moved by more than $10$ lattice sites.
However, the situation is more fortunate, since subsequent steps are correlated, as shown in Fig.~\ref{fig:stability}(b).
Once this correlation is taken into account, we estimate an RMS drift of 8(1)~sites over 1000~realizations in the transverse direction and 12(2)~sites over 1000~realizations in the axial direction.
This drift corresponds to an RMS displacement of $\unit{14(3)\times 10^{-3}}{sites / realization}$, compared to a beam waist of $\approx$\unit{80}{\mu{}m} ($\approx$140 sites).

@@

In general, we cannot extract the lattice position from the atomic distribution directly, but rely on the monitor camera picture, which is related to the lattice position in the atom plane by an affine transformation, consisting of a rotation, a scaling, and an offset.
To determine the parameters of this affine transform, we use the remote actuators to purposely misalign the lattice beams and correlate the monitor camera image with the atom image obtained by fluorescence imaging~\cite{Parsons2015}.
The corresponding least-squares fit is shown in Fig.~\ref{fig:stability}(c).
The quality of this fit remains acceptable for $\approx$30 repetitions of the experiment, but it degrades for longer times.
This effect is most likely due to thermal drift of the beam-monitoring camera, since a displacement of a one pixel (\unit{3}{\mu{}m}) on the monitoring camera is equivalent to a displacement of one lattice site in the atom plane.
The most likely parameter of the affine transform susceptible to thermal drift over a few experimental cycles is the offset.

@@

In conclusion: the beam-monitoring camera is sufficiently sensitive to detect lattice misalignments on the order of a few lattice sites, making it useful as a noninvasive check of the lattice position.

>>>>

\FIXME{}: begin Sebastian's version
{\color{red}
Variation in external conditions can contaminate large datasets if the prepared state sensitively depends on laser pointing.
To guard against such contamination, we have implemented an independent monitoring system to verify appropriate pointing of both lattice laser beams.
This monitoring system allows us to eliminate (via post-selection) experimental realizations during which the system was in an undesired state.
A beam-sampler (Thorlabs BSF10-C) reflects a small portion of the beam shortly before it reaches the atoms and images it onto a monitoring camera located in the plane corresponding to the atomic system, imaging a Gaussian beam corresponding to the optical lattice.
Naturally, the position of the atoms is dictated by the light field of both the incoming and retroreflected beams, while the monitoring camera only observes the position of the incoming beam.

@@

We first characterize the passive stability of the optical lattice along one dimension by monitoring its position over the course of 250 experimental realizations.
We decompose the lattice position into the transverse $x_t$ and the longitudinal $x_l$ components with respect to the optical axis.
Figure~\ref{fig:stability}(a) shows a histogram of displacements of atoms between subsequent experimental realizations, showing that the RMS displacement is $0.54(3)$ [$0.98(4)$] sites along the $x_t$ [$x_l$] direction.
The variation of the displacements is a geometric effect consistent with the mechanical assembly of the experiment~\cite{Huber2014}.
The distributions are centered within 1-$\sigma$ of zero along both directions: $-0.03(4)$ and $0.01(5)$ sites for transverse and longitudinal axes, respectively.
If the trap had followed a random walk with these step sizes, the optical lattice would have to be realigned frequently, since after only $100$ experimental shots, the trap would have moved by more than $10$ lattice sites.
However, the situation is more fortunate, since subsequent steps are correlated, as shown in Fig.~\ref{fig:stability}(b).
Once this correlation is taken into account, we estimate an RMS drift of 8(1)~sites over 1000~realizations in the transverse direction and 12(2)~sites over 1000~realizations in the axial direction.
This drift corresponds to an RMS displacement of $\unit{14(3)\times 10^{-3}}{sites / realization}$, compared to a beam waist of $\approx$\unit{80}{\mu{}m} ($\approx$140 sites).
}
\FIXME{}: Begin Daniel's version
{\color{green}
We characterized the passive stability of the optical lattice by monitoring the position over the course of 250 sequential experimental realizations.
To analyse the data we decomposed the recorded positions into the transverse $x_t$ and longitudinal $x_l$ components as compared to the optical axis (i.e. rotated by $45^\circ$ as compared to the lattice directions $x$ and $y$).
During this measurement we found that the drift of the atom position is about $1-2$ lattice site, see red data in Fig.~\ref{fig:stability}(e+f).
For a more in-depth analysis, we also studied the histogram of displacements of atoms between subsequent experimental realizations, see Fig.~\ref{fig:stability}(a).
This shows a RMS displacement of $0.54(3)$ [$0.98(4)$] sites along the $x_t$ [$x_l$] direction.
The difference between the displacements is a geometric consequence of the mechanical assembly of the experiment~\cite{Huber2014}.
The distributions are centered within 1-$\sigma$ of zero along both directions: $-0.03(4)$ and $0.01(5)$ sites for transverse and longitudinal axes, respectively, indicating the absence of strong drift.
If the trap followed a random walk with these step sizes, the optical assembly would have to be frequently realigned, since after only $100$ experimental shots the trap would move by more than $10$ lattice sites.
The situation is more fortunate, however, since subsequent steps are correlated, as shown in Fig.~\ref{fig:stability}(b).
Once this correlation is taken into account, we estimate an RMS drift of \unit{8(1)} [12(2)] {sites/ 1000 realizations} in the transverse [axial] direction, compared to a beam waist of $\approx$140 sites.
}

@@

\FIXME{}: Begin Sebastian's version
{\color{red}
In general, we cannot extract the lattice position from the atomic distribution directly, but rely on the monitor camera picture, which is related to the lattice position in the atom plane by an affine transformation, consisting of a rotation, a scaling, and an offset.
To determine the parameters of this affine transform, we use the remote actuators to purposely misalign the lattice beams and correlate the monitor camera image with the atom image obtained by fluorescence imaging~\cite{Parsons2015}.
The corresponding least-squares fit is shown in Fig.~\ref{fig:stability}(c).
The quality of this fit remains acceptable for $\approx$30 repetitions of the experiment, but it degrades for longer times.
This effect is most likely due to thermal drift of the beam-monitoring camera, since a displacement of a one pixel (\unit{3}{\mu{}m}) on the monitoring camera is equivalent to a displacement of one lattice site in the atom plane.
The most likely parameter of the affine transform susceptible to thermal drift over a few experimental cycles is the offset.

@@

In conclusion: the beam-monitoring camera is sufficiently sensitive to detect lattice misalignments on the order of a few lattice sites, making it useful as a noninvasive check of the lattice position.
}
\FIXME{}: Begin Daniel's version
{\color{green}
As an additional measure, we have implemented a monitoring system to independently verify appropriate pointing of the lattice laser beams.
This is particularly useful when performing experiments where imaging the atoms does note allow determining the original center of the atoms in the harmonic trap.
Additionally, this beam-monitoring system could allow us to eliminate (via post-selection) experimental realizations when the system was in an undesired state -- or designing a feedback loop on the lattice beam pointing to ensure long-term stability.
A beam-sampler (Thorlabs BSF10-C) reflects a small portion of the beam shortly before it reaches the atoms and images it onto a camera located in the plane corresponding to the atomic system, imaging a Gaussian beam identical to that of the optical lattice.
Naturally, the position of the atoms is dictated by the light field of both the incoming and retro-reflected beams, while the monitoring camera only observes the position of the in-going beam (this is found to not be a limitation in practice).

As a first step, we calibrated the beam-monitoring system by simultaneously recording the position of the atoms on the imaging camera (Andor iStar) and the lattice beam position on the beam-monitoring camera (Point Grey Chameleon).
By additionally varying the laser beam position and the lattice anisotropy we were able to obtain a reliable calibration, see Fig.~\ref{fig:stability} (c).
However, we found that this calibration changes after a short while ($\approx$ 30 shots) and there is an offset between the recordings of the beam-monitoring camera and the atoms camera.
This is most likely caused by thermal drifts of the beam-monitoring camera housing: the camera heats up by a few $^\circ$C owing to the absence of active cooling. This can easily lead to a few $\mu$m drift due to the plastic mounting between the post holder mounted to the optical table and the CCD chip.
While this could be easily fixed with an improved camera design, we implemented an alternative approach: by recalibrating the offset every 20 shots, this effect can be taken into account.
After correcting for this changing offset of the beam-monitoring camera, we find very good agreement between the recorded atom position with the imaging camera and the predicted atom position based on the beam-monitoring camera, see Fig.~\ref{fig:stability}(d-f).
The average $L_1$ displacement between the predicted position and measured position is 0.7~lattice sites.
The beam-monitoring system is thus sufficiently sensitive to detect lattice misalignments that are on the order of a few lattice sites or more, making it useful as a non-invasive check of the alignment of the experimental apparatus.
}
